---
title: "Project1"
output: html_document
date: "2025-10-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library(NHANES)
library(ggcorrplot)
library(ggplot2)
library(dplyr)
library(tidyverse)
```

# Draft Set of Questions

# 1. How does AlcoholDay compare between Married and Divorced marital status groups across Race3

```{r}
nhanes_5 <- NHANESraw |>
  filter(SurveyYr == "2011_12" & Age >= 20)|>
  mutate(MaritalStatus2 = case_when(
    MaritalStatus %in% c("NeverMarried", "Divorced") ~ "Not Married",
    MaritalStatus %in% c("Married") ~ "Married"),
    MaritalStatus2 = factor(MaritalStatus2, levels = c("Married", "Not Married")))|>
  mutate(LogAlc = log(AlcoholYear))|>
  filter(!is.na(MaritalStatus2), 
         !is.na(Race3),
         !is.na(LogAlc))|>
  select(MaritalStatus2, Race3, LogAlc, AlcoholYear)

ggplot(nhanes_5,aes(x = LogAlc, fill = MaritalStatus2)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~Race3) + 
  labs(title = "Distribution of Estimated Number of Days Per Year Participants Drank
       Alcohol by Race and Marriage Status", 
       x = "Estimated Number of Days Drinking Per Year (log)", 
       y = "Density", fill = "Marital Status",
       caption = str_wrap("This graph shows the KDE plots of the log-transformed self-reported consumption of alcoholic beverages in the past year for participants across the Race 3 variable for the 2011-2012 survey year. Log transformation was applied to normalize the distribution and improve visualization clarity by reducing the influence of high outliers. Only participants 20+ years old were included. Marital status is split into two variables: 'Married' and 'Not Married'. The 'Asian', 'Mexican' and 'Other' racial group distributions are bimodal, while 'Hispanic', 'White' and 'Black' racial groups are multimodal. Source: NHANESraw dataset", width = 100)) +  
  theme(plot.title = element_text(size = 12, hjust = 0.8))+
  theme_bw()
```
```{r}
# Calculate summary statistics
summary_stats<-nhanes_5 |>
  group_by(Race3, MaritalStatus2) |>
  summarise(
    mean_days = mean(AlcoholYear),
    median_days = median(AlcoholYear),
    se = sd(AlcoholYear) / sqrt(n()),
    LCI = mean_days - se,
    UCI = mean_days + se
  )

# Plot means with error bars
ggplot(summary_stats, aes(x = Race3, y = mean_days, fill = MaritalStatus2)) +
  geom_col(position = position_dodge(0.9), alpha = 0.7) +
  geom_errorbar(aes(ymin = LCI, ymax = UCI),
                position = position_dodge(0.8), width = 0.2) +
  geom_point(aes(y = median_days, shape = MaritalStatus2), 
             position = position_dodge(0.8), size = 2, color = "black") +
  labs(title = "Mean Number of Days Consuming Alcoholic Beverages in the Past Year by Race vs Median",
       x = "Race", y = "Drinking Days per Year", fill = "Drinking Days Per Year (Mean)", shape = "Drinking Days Per Year (Median)", 
       caption = str_wrap( "This graph compares the means and medians of the self-reported amount of alcoholic beverages consumed per year across the Race3 variable. The bars show means (with 95% CIs) and the shapes show medians for drinking days per year. The 'White' race category exhibits nearly identical median drinking days despite differences in the mean, whereas the 'Other' category shows substantial variation in both the mean and median, with differences in the median being significantly higher. Source: NHANESraw dataset"), width = 200)+
  guides(fill = guide_legend(override.aes = list(shape = NA))) +
  theme(plot.title = element_text(hjust = 0.2))
  theme_bw()
```
```{r}
pie<- nhanes_5|>
  count(Race3,MaritalStatus2, name = "count")|>
  group_by(Race3)|>
  mutate(percentage = round(count / sum(count)*100, digits = 1))|>
  ungroup()
ggplot(pie, aes(x = "", y = count, fill = MaritalStatus2)) + 
    geom_bar(stat = "identity", position = position_fill(), 
             width = 1, color = "white") +
    geom_text(aes(label = count), position = position_fill(vjust = 0.5), size = 3.5) +
    geom_text(aes(label = paste0("\n","\n","(",percentage,"%)")), position = position_fill(vjust = 0.5), size = 2.4)+
    coord_polar("y", start = 0)+
    facet_wrap(~ Race3)+labs(title = "Marital Status by Race", 
                             fill = "Marital Status", 
                             caption = str_wrap("This is a faceted pie chart showing the distribution of marital status across the Race3 variable. Each pie represents one race category, with slices corresponding to one of two marital status groups. The size of each slice reflects the proportion of individuals within that race who fall into each marital status category, with counts both the number of participants in each group and the percentage of the whole. The 'Other' race category is evenly split among 'Married' and 'Not Married', but also has the fewest respondents. The 'Mexican' category has the highest ratio of 'Married' to 'Not Married', while 'Black' has the lowest. Source: NHANESraw dataset", width = 100))+ 
  theme(plot.title = element_text(hjust = 0.5))+
  theme_void()+theme(plot.title = element_text(hjust = 0.5))
```

# 2. What is the distribution of testosterone in the data set over age and how does it vary across BMI groups?

## Testosterone by Age Across BMI WHO Groups for Males

-   Testosterone variable is total testosterone in nanograms per deciliter for participants aged 6 years or older for the years 2011-2012.
-   Age is the age of the participant in 2011-2012, subjects who were 80 years or older were recorded as 80
-   BMI WHO is the body mass index as categorized by the World Health Organization as 12.0-18.4 (underweight), 18.5-24.9 (normal weight), 25-29.9 (overweight), and 30 plus (Obese).

```{r}
NHANES_t <- NHANESraw |>
  select('Testosterone', 'Age', 'Gender','BMI_WHO') |>
  mutate(BMI_WHO = case_when(
    BMI_WHO == '12.0_18.5' ~ 'Underweight',
    BMI_WHO == '18.5_to_24.9' ~ 'Normal Weight',
    BMI_WHO == '25.0_to_29.9' ~ 'Overweight',
    BMI_WHO == '30.0_plus' ~ 'Obese'
  ))

NHANES_t |>
  filter(Gender == 'male') |>
  drop_na(BMI_WHO)|>
  ggplot(aes(x = Age, y = Testosterone)) +
  geom_point(aes(col = BMI_WHO), size = 0.65)+
  facet_wrap(~BMI_WHO) +
  labs(title = 'Testosterone vs Age', y = 'Testosterone (ng/dL)',
       col = 'BMI WHO', caption = 'Source: NHANESraw dataset') + 
  theme_bw()
```

-   Since the distributions are not normal, robust statistics are appropriate for better understanding the distributions.
  - Median and interquartile range are used to accurately compare the spread of testosterone data by age group overtime.

## Median Testosterone by Age Across BMI WHO groups.

```{r}
NHANES_t2 <- NHANES_t |>
  filter(Gender == 'male')|>
  drop_na(BMI_WHO, Testosterone)|>
  mutate(age_group = case_when(
    Age <= 20 ~ '06-20',
    Age <= 40 ~ '21-40',
    Age <= 60 ~ '41-60',
    Age <= 80 ~ '61-80'
  ))|>
  group_by(age_group, BMI_WHO) |> 
  summarize(medT = median(Testosterone), Q1 = quantile(Testosterone, 0.25), Q3 = quantile(Testosterone, 0.75))

ggplot(NHANES_t2, aes(x = age_group, y = medT))+
  geom_bar(aes(fill = BMI_WHO), stat = 'identity', position = 'dodge') + 
  geom_errorbar(aes(ymin = Q1, ymax = Q3, group = BMI_WHO),
                position=position_dodge(0.9))+
  geom_point(aes(group = BMI_WHO), position = position_dodge(0.9))+
  labs(title = 'Testosterone vs Age Group',
       y = 'Median Testosterone', x = 'Age Group',fill = 'BMI WHO',
       caption = 'Source: NHANESraw Dataset, Error Bar: First Quartile, Median, Third Quartile') + 
  theme_bw()

```
- Age is grouped into approximately twenty year increments starting at the first age recorded for testosterone which is six years

# 3. How does poverty ratio relate to BMI and cholesterol levels across races?
```{r}
# Filter to adults in 2011_12 with complete data for variables of interest.
nhanes_clean <- NHANESraw |> 
  filter(SurveyYr == "2011_12", Age >= 18) |> 
  select(ID, Race1, Race3, Poverty, BMI, TotChol, Age, Gender) |> 
  na.omit()
```
# Poverty vs BMI by Race
```{r, fig.width=8}
# Race3 distribution
ggplot(nhanes_clean, aes(x = Poverty, y = BMI)) +
  geom_point(alpha = 0.5, aes(color = Race3), size = 0.6) +
  geom_smooth(se = TRUE, color = "black") +
  facet_wrap(~ Race3) +
  labs(title = "Relationship Between Poverty Ratio and BMI Across Race",
    x = "Poverty Ratio", y = "Body Mass Index (BMI)", 
    caption = str_wrap("Source: NHANESraw 2011–2012 (adults 18+). This plot shows how poverty ratio (income relative to the poverty line) relates to BMI across racial groups. Most participants fall between poverty ratios of 0–2 and BMI values of 20–40. Trend lines are mostly flat, showing little association between income and BMI. The confidence intervals are narrow for most groups, indicating consistent patterns, but noticeably wider for the “Other” category due to smaller sample size.", width=100)
  ) 
```
# Poverty vs Cholesterol by Race
```{r, fig.width=8}
# Race3 distribution
ggplot(nhanes_clean, aes(x = Race3, y = TotChol, color = Race3)) +
  geom_jitter(alpha = 0.3, width = 0.2, size = 0.6) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  labs(
    title = "Total Cholesterol by Race and Poverty Ratio Groups",
    x = "Race Category", y = "Total Cholesterol (mmol/L)", 
       caption = str_wrap("Source: NHANESraw 2011–2012 (adults 18+). Distribution of total cholesterol (mmol/L) across racial groups. Values cluster around 5 mmol/L, with small variations in median and spread. White and Black groups have denser samples, while Hispanic participants show greater variability.", width=110)
  )
```

# 4. To what extent is education associated with the age when participants first begin smoking or using marijuana regularly, and does this vary by gender?
```{r, fig.width=8}
# Filtering data 
nhanes_clean2 <- NHANESraw |> 
  select(ID, Age, Gender, SmokeAge, AgeRegMarij, Education) |> 
  na.omit() |>
  mutate(
    SmokeAgeGroup = cut(SmokeAge, breaks=c(0,14,17,20,100), labels=c("<15","15-17","18-20","21+")),
    MarijAgeGroup = cut(AgeRegMarij, breaks=c(0,14,17,20,100), labels=c("<15","15-17","18-20","21+"),           include.lowest = TRUE)
  )

# Smoking vs Education
ggplot(nhanes_clean2, aes(x=Education, fill=SmokeAgeGroup)) +
  geom_bar(position="dodge") + facet_wrap(~Gender) +
  labs(title="Age Started Smoking by Education and Gender",
       x="Education Level", y="Number of Participants", fill="Age Group", 
       caption = str_wrap("Source: NHANESraw. Age participants began smoking regularly, grouped by education level and gender. Across both males and females, most started smoking between ages 15–17. The highest counts appear among those with a high school or some college education. College graduates also show smoking initiation around 15–17, but in smaller numbers. Males generally have higher counts, likely due to a larger sample size.", width=80)) + coord_flip()
```

# Marijuana vs Education faceted by Gender
```{r, fig.width=8}
ggplot(nhanes_clean2, aes(x=Education, fill=MarijAgeGroup)) +
  geom_bar(position="dodge") + facet_wrap(~Gender) +
  labs(title="Age Started Marijuana Use by Education and Gender",
       x="Education Level", y="Number of Participants", fill="Age Group", 
       caption = str_wrap("Source: NHANESraw. Age participants began using marijuana, grouped by education level and separated by gender. Similar to previous graph where most participants begin using between ages 15 and 17, with the highest counts among those with a high school education or some college. Male participants generally have higher counts, particularly among those in high school, while females show the highest counts in the 15–17 age group for both high school and some college. For college graduates, initiation is slightly later, peaking around 18–20.", width=80)) + coord_flip()
```
